---
title: "Downloading and Visualizing Camels Data"
author: "Matthew Ross"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(GGally)

knitr::opts_chunk$set(echo = TRUE)
```


# Data Acquisition

For this assignment we are going to be playing with annually
aggregated metrics of USGS flow data from the [CAMELS](https://ral.ucar.edu/solutions/products/camels) dataset. This dataset
has sparked a revolution in machine learning in hydrology. 


```{r}
if(!file.exists('data')){
  dir.create('data')
}

#climate means
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_clim.txt',
              'data/climate.txt')

#geologic controls
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_geol.txt',
              'data/geol.txt')

# soil data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_soil.txt',
              'data/soil.txt')

# topo data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_topo.txt',
              'data/topo.txt')


#vege 

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_vege.txt',
              'data/vege.txt')


#hydro (response variable)

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_hydro.txt',
              'data/hydro.txt')


# Variable definitions
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf',
              'data/meta.pdf')

```


## Data org


```{r}
dat_files <- list.files('data',
                        full.names = T)



climate <- read_delim(dat_files[1], delim = ';')

hydro <- read_delim('data/hydro.txt', delim = ';')

geol <- read_delim('data/geol.txt', delim = ';')

soil <- read_delim('data/soil.txt', delim = ';')

topo <- read_delim('data/topo.txt', delim = ';')

vege <- read_delim('data/vege.txt', delim = ';')

```

## Initial data viz


### Baseflow 


```{r}

ggplot(hydro, aes(x = baseflow_index,
                  y = q95)) + 
  geom_point()


```


Baseflow doesn't strongly control Q95 in a predictable way. 



### Climate controls


```{r}

cq <- inner_join(climate, hydro %>%
                   select(gauge_id, q95))



ggplot(cq, aes(x = p_mean, y = q95)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

p_mean_mod <- lm(q95 ~ p_mean, data = cq)

```


#### All at once

```{r}

png(filename = 'bigclimeplot.png', width = 10, height = 8, units = 'in', res = 300)
cq %>%
  select_if(is.numeric) %>%
  ggpairs()
dev.off()


long_cq <- cq %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = 'value',
               names_to = 'driver')

ggplot(long_cq, aes(value,
                    q95)) +
  geom_point() +
  facet_grid(~driver,
             scales = 'free')
```

The average precip (p_mean) controls 71% of the variation in 195, where every 1 mm/day increase in long-term average precip increases the q95 by 2.95 mm/day. 

## Assignment code


# Assignment

## What are three controls on average runoff ratio? 

```{r}
hydro_soil<- inner_join(hydro%>%
                   select(gauge_id, runoff_ratio, baseflow_index, q_mean),soil)

ggplot(hydro_soil, aes(x = clay_frac, y = runoff_ratio)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

runoff_clay_mod <- lm(runoff_ratio ~ clay_frac, data = hydro_soil)

summary(runoff_clay_mod)
```

My first predictor clay fraction is statistically significant with an extremely low p-value (< 2.2e-16), however it has a low explainability as a driver with only a 10.56 % r2 value. 

```{r}

ggplot(hydro_soil, aes(x = soil_porosity, y = runoff_ratio)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

runoff_porosity_mod <- lm(runoff_ratio ~ soil_porosity, data = hydro_soil)

summary(runoff_porosity_mod)
```

Soil porisity as another predictor is also statistically significant (p-val 0.00633) but with an almost negligible r2 value, of less than 1%, in this case I would choose clay fraction as a stronger driver with what I would consider a similar contribution of porosity. 

```{r}
hydro_topo<- inner_join(hydro%>%
                   select(gauge_id, runoff_ratio, baseflow_index, q_mean),topo)

ggplot(hydro_topo, aes(x = slope_mean, y = runoff_ratio)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

runoff_water_mod <- lm(runoff_ratio ~ slope_mean, data = hydro_topo)

summary(runoff_water_mod)
```

For the last contol of runoff ratio I used the topographical data instead of soil and slope mean has both a strong p value (<2e-16), and explainability (~38%). Additionally, I think this would combine with clay fraction as strong co-predictors in a ML/DL model becasue it is a different type of trend the model can pick up on. 

## What are three controls on baseflow_index? 
```{r}
ggplot(hydro_soil, aes(x = clay_frac, y = baseflow_index)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

baseflow_clay_mod <- lm(baseflow_index ~ clay_frac, data = hydro_soil)

summary(baseflow_clay_mod)
```

Clay fraction is similary to runoff in both statistical significance (p_val <2e-16) and explainability (~14%), it is a driver but not majorly which makes sense since higher clay fractions would allow for lower infiltration with fast precip events. 

```{r}
ggplot(hydro_soil, aes(x = sand_frac, y = baseflow_index)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

baseflow_sand_mod <- lm(baseflow_index ~ sand_frac, data = hydro_soil)

summary(baseflow_sand_mod)
```

I chose to model sand fraction becasue I thought it would lead to more baseflow since there is a higher porosity but  in this case it is less significant both statistically (p-val = 1.01e-14 ) and explainability (8%) than clay fraction. Between the two I would go with clay fraction. 

```{r}
hydro_climate<- inner_join(hydro%>%
                   select(gauge_id, runoff_ratio, baseflow_index, baseflow_index, q_mean),climate)

ggplot(hydro_climate, aes(x = frac_snow, y = baseflow_index)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

baseflow_snow_mod <- lm(baseflow_index ~ frac_snow, data = hydro_climate)

summary(baseflow_snow_mod)
```

Lastly for baseflow I chose fraction snow from the climate data to compare to because i thought that the snow dominated hydrogrpahs would have a stronger relationship with baseflow as the snow should melt slower and allow for more groundwater inputs downstream. P val (<2e-16 ) and r2 (15%) makes it as strong as clay fraction, there may be some overlap if using them as coverinece. 

## What are three controls on mean flow? 
```{r}
png(filename = 'qmean_soil.png', width = 10, height = 8, units = 'in', res = 300)
hydro_soil %>%
  select_if(is.numeric) %>%
  ggpairs()
dev.off()
```

with respect to the soil variables being a driver of average discharge, I would likely elect to not use any of the soil variables for the prediction of q_mean because there doesn't seem to be a high correlation between any of the variables and q_mean. 

```{r}
png(filename = 'qmean_climate.png', width = 10, height = 8, units = 'in', res = 300)
hydro_climate %>%
  select_if(is.numeric) %>%
  ggpairs()
dev.off()
```

A lot of the climate variables seem to be highly correlated with mean discharge which makes sense since these account for precipitation and intensity, I would likely choose two or three of the variables that may not have a lot of covariance for a model predicting mean discharge, namely; aridity, p_mean, and either the paring of high/low precip duration/frequency, I think that aridity would be correlated wih p_mean. 

```{r}
png(filename = 'qmean_topo.png', width = 10, height = 8, units = 'in', res = 300)
hydro_topo %>%
  select_if(is.numeric) %>%
  ggpairs()
dev.off()
```

Lastly, for topographical modeling, I would likely choose the variables gauge latitude and slope mean, I would choose gauge latitude because it makes sense that lower latitudes would have lower discharges since they are more arid. I also think that slope mean would give a model more context as I think that more gradual slopes would have higher mean discharges since they would likely be located in higher order streams (downstream). 