---
title: "Downloading and Visualizing Camels Data"
author: "Matthew Ross"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(GGally)

knitr::opts_chunk$set(echo = TRUE)
library(GGally)
```

# Data Acquisition

For this assignment we are going to be playing with annually aggregated metrics of USGS flow data from the [CAMELS](https://ral.ucar.edu/solutions/products/camels) dataset. This dataset has sparked a revolution in machine learning in hydrology.

```{r}
if(!file.exists('data')){
  dir.create('data')
}

#climate means
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_clim.txt',
              'data/climate.txt')

#geologic controls
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_geol.txt',
              'data/geol.txt')

# soil data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_soil.txt',
              'data/soil.txt')

# topo data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_topo.txt',
              'data/topo.txt')


#vege 

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_vege.txt',
              'data/vege.txt')


#hydro (response variable)

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_hydro.txt',
              'data/hydro.txt')


# Variable definitions
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf',
              'data/meta.pdf')

```

## Data org

```{r}
dat_files <- list.files('data',
                        full.names = T)



climate <- read_delim(dat_files[1], delim = ';')

hydro <- read_delim('data/hydro.txt', delim = ';')

soil <- read_delim('data/soil.txt', delim = ';')

geol <- read_delim('data/geol.txt', delim = ';')
```

## Initial data viz

### Baseflow

```{r}

#hydroclimate
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, q95))
#baseflow 
ggplot(hydro, aes(x=baseflow_index, y=q95))+
  geom_point()


ggplot(climate_q, aes(x=p_mean, y=q95))+
  geom_point()+
  geom_smooth(method = "lm", se=F)

p_mean_mod <- lm(q95~p_mean, data=climate_q)
summary(p_mean_mod)
```

\*r\^2 \> 0.3

correlation ^2^ = r\^2

```{r}

ggplot(hydro, aes(x = baseflow_index,y = q95)) + 
  geom_point()

```

Baseflow doesn't strongly control Q95 in a predictable way.

### Climate controls

```{r}

cq <- inner_join(climate, hydro %>%
                   select(gauge_id, q95))



ggplot(cq, aes(x = p_mean, y = q95)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

p_mean_mod <- lm(q95 ~ p_mean, data = cq)

```

#### All at once

```{r}

png(filename = 'bigclimeplot.png', width = 10, height = 8, units = 'in', res = 300)
cq %>%
  select_if(is.numeric) %>%
  ggpairs()
dev.off()


long_cq <- cq %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = 'value',
               names_to = 'driver')

ggplot(long_cq, aes(value,
                    q95)) +
  geom_point() +
  facet_grid(~driver,
             scales = 'free')
```

The average precip (p_mean) controls 71% of the variation in 195, where every 1 mm/day increase in long-term average precip increases the q95 by 2.95 mm/day.

# Assignment

## What are three controls on average runoff ratio?

```{r}
#climate runoff ratio join
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, runoff_ratio))

#climate x hydro runoff ratio correlatios plots
png(filename = "bigclimate.png", width = 10, height = 8, units = "in", res = 300)
climate_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#runoff ratio x aridity

ggplot(climate_q, aes(x=runoff_ratio, y=aridity))+
  geom_point()

#lm rr x aridity
climate_q_mod <- lm(runoff_ratio~aridity, data=climate_q) 
summary(climate_q_mod) 

 
```

r\^2 = .327

The aridity factor (aridity) explains 32.84% of the variability in runoff ratio, with every unit increase in aridity decreasing the runoff ratio by 0.21707, as determined by the linear regression model (p \< 2e-16).

```{r}
#runoff ratio x low_precip_freq
ggplot(climate_q, aes(x=runoff_ratio, y= low_prec_freq))+
  geom_point() 

#lm rr x low precip freq
climate_q_mod <- lm(runoff_ratio~low_prec_freq, data=climate_q) 
summary(climate_q_mod) 

```

r\^2 =.527

The frequency of low precipitation events (low_prec_freq) accounts for 52.69% of the variability in runoff ratio, with every increase in the frequency of low precipitation events by one unit leading to a decrease in the runoff ratio by 0.0048183, as determined by the linear regression model (p \< 2e-16).

```{r}
#runoff ratio x high_precip_freq
ggplot(climate_q, aes(x=runoff_ratio, y= high_prec_freq))+
  geom_point() 

#lm rr x high precip freq
climate_q_mod <- lm(runoff_ratio~high_prec_freq, data=climate_q) 
summary(climate_q_mod) 
```

r\^2 = .462

The frequency of high precipitation events (high_prec_freq) explains 46.27% of the variability in runoff ratio, with each unit increase in the frequency of high precipitation events leading to a decrease in the runoff ratio by 0.034879, as determined by the linear regression model (p \< 2e-16).

## What are three controls on baseflow_index?

```{r}
#geol baseflow join
geol_q <- inner_join(geol, hydro %>%
                          select(gauge_id, baseflow_index))

#geol x hydro runoff ratio correlatios plots
png(filename = "biggeol.png", width = 10, height = 8, units = "in", res = 300)
geol_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#soil baseflow join
soil_q <- inner_join(soil, hydro %>%
                          select(gauge_id, baseflow_index))

#soil x hydro runoff ratio correlatios plots
png(filename = "bigsoil.png", width = 10, height = 8, units = "in", res = 300)
soil_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#climate runoff ratio join
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, baseflow_index))

#climate x hydro runoff ratio correlatios plots
png(filename = "bigclimate.png", width = 10, height = 8, units = "in", res = 300)
climate_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#hydro x hydro
png(filename = "bighydro.png", width = 10, height = 8, units = "in", res = 300)
hydro %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 


#baseflow index x low discharge frequency
ggplot(hydro, aes(x=baseflow_index, y= low_q_freq))+
  geom_point() 

hydro_mod <- lm(baseflow_index~low_q_freq, data=hydro) 
summary(hydro_mod) 
```

r\^2 = .738

The frequency of low streamflow events (low_q_freq) explains 73.81% of the variability in baseflow index, with each increase in the frequency of low streamflow events by one unit resulting in a decrease in the baseflow index by 0.001697, as determined by the linear regression model (p \< 2e-16).

```{r}
#baseflow index x high discharge frequency
ggplot(hydro, aes(x=baseflow_index, y= high_q_freq))+
  geom_point() 

hydro_mod <- lm(baseflow_index~high_q_freq, data=hydro) 
summary(hydro_mod) 
```

r\^2 = .397

The frequency of high streamflow events (high_q_freq) accounts for 39.72% of the variability in baseflow index, with each increase in the frequency of high streamflow events by one unit resulting in a decrease in the baseflow index by 0.0035214, as determined by the linear regression model (p \< 2e-16).

```{r}
#baseflow index x runoff ratio
ggplot(hydro, aes(x=baseflow_index, y= runoff_ratio))+
  geom_point() 

hydro_mod <- lm(baseflow_index~runoff_ratio, data=hydro) 
summary(hydro_mod) 
```

The runoff ratio explains 11.92% of the variability in baseflow index, with each unit increase in runoff ratio leading to an increase in the baseflow index by 0.24046, as determined by the linear regression model (p \< 2e-16).

## What are three controls on mean flow?

```{r}
#climate x mean flow 
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, q_mean))

#climate x hydro mean flow correlatios plots
png(filename = "bigclimate.png", width = 10, height = 8, units = "in", res = 300)
climate_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#mean flow x aridity
ggplot(climate_q, aes(x=q_mean, y= aridity))+
  geom_point() 

hydro_mod <- lm(aridity~q_mean, data=climate_q) 
summary(hydro_mod) 

```

r\^2= .338

The mean flow (q_mean) accounts for 33.85% of the variability in aridity, with every unit increase in mean flow resulting in a decrease in aridity by 0.23217, as determined by the linear regression model (p \< 2e-16).

```{r}
#mean flow x mean precip
ggplot(climate_q, aes(x=q_mean, y= p_mean))+
  geom_point() 

hydro_mod <- lm(p_mean~q_mean, data=climate_q) 
summary(hydro_mod) 
```

r\^=.786

The mean flow (q_mean) explains 78.5% of the variability in mean precipitation (p_mean), with every unit increase in mean flow leading to an increase in mean precipitation by 0.81095, as determined by the linear regression model (p \< 2e-16).

```{r}
#mean flow x high precip freq
ggplot(climate_q, aes(x=q_mean, y= high_prec_freq))+
  geom_point()  

hydro_mod <- lm(high_prec_freq~q_mean, data=climate_q) 
summary(hydro_mod) 
```

r\^2 = .4472

The mean flow (q_mean) explains 44.72% of the variability in the frequency of high precipitation events (high_prec_freq), with every unit increase in mean flow leading to a decrease in the frequency of high precipitation events by 1.9714, as determined by the linear regression model (p \< 2e-16).
