---
title: "Downloading and Visualizing Camels Data"
author: "Matthew Ross"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(GGally)

knitr::opts_chunk$set(echo = TRUE)
```

# Data Acquisition

For this assignment we are going to be playing with annually aggregated metrics of USGS flow data from the [CAMELS](https://ral.ucar.edu/solutions/products/camels) dataset. This dataset has sparked a revolution in machine learning in hydrology.

```{r}
if(!file.exists('data')){
  dir.create('data')
}

#climate means
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_clim.txt',
              'data/climate.txt')

#geologic controls
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_geol.txt',
              'data/geol.txt')

# soil data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_soil.txt',
              'data/soil.txt')

# topo data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_topo.txt',
              'data/topo.txt')


#vege 

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_vege.txt',
              'data/vege.txt')


#hydro (response variable)

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_hydro.txt',
              'data/hydro.txt')


# Variable definitions
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf',
              'data/meta.pdf')

```

## Data org

```{r}
dat_files <- list.files('data',
                        full.names = T)



climate <- read_delim(dat_files[1], delim = ';')

hydro <- read_delim('data/hydro.txt', delim = ';')

```

## Initial data viz

### Baseflow

```{r}

ggplot(hydro, aes(x = baseflow_index,
                  y = q95)) + 
  geom_point()
```

Baseflow doesn't strongly control Q95 in a predictable way.

### Climate controls

```{r}

cq <- inner_join(climate, hydro %>%
                   select(gauge_id, q95))


ggplot(cq, aes(x = p_mean, y = q95)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

p_mean_mod <- lm(q95 ~ p_mean, data = cq)
```

#### All at once

```{r}

png(filename = 'bigclimeplot.png', width = 10, height = 8, units = 'in', res = 300)

cq %>%
  select_if(is.numeric) %>%
  ggpairs()

dev.off()


long_cq <- cq %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = 'value',
               names_to = 'driver')

ggplot(long_cq, aes(value,
                    q95)) +
  geom_point() +
  facet_grid(~driver,
             scales = 'free')
```

The average precip (p_mean) controls 71% of the variation in 195, where every 1 mm/day increase in long-term average precip increases the q95 by 2.95 mm/day.

# Assignment

## What are three controls on average runoff ratio?

```{r}
cq_rr <- inner_join(climate, hydro %>%
  select(gauge_id, runoff_ratio))
```

```{r}
png(filename = 'bigrrplot.png', width = 10, height = 8, units = 'in', res = 300)

cq_rr %>%
  select_if(is.numeric) %>%
  ggpairs()

dev.off()
```

```{r}
#plot relationships
ggplot(cq_rr, aes(x=p_mean, y = runoff_ratio)) + geom_point() + geom_smooth(method = "lm", se = F)

ggplot(cq_rr, aes(x=high_prec_freq, y = runoff_ratio)) + geom_point() + geom_smooth(method = "lm", se = F)

ggplot(cq_rr, aes(x=low_prec_freq, y = runoff_ratio)) + geom_point() + geom_smooth(method = "lm", se = F)
```

```{r}
#model relationships to get summarys

p_mean_rr_mod <- lm(runoff_ratio~p_mean, data = cq_rr)

summary(p_mean_rr_mod)

##repeat this for the other two! Then write a description of what we found
```

Three controls on average runoff ratio include p_mean, high precipitation frequency, and low precipitation duration

## What are three controls on baseflow_index?

```{r}
cq_bf <- inner_join(geol, hydro %>%
                      select(gauge_id, baseflow_index))
```

```{r}

```

## What are three controls on mean flow?

```{r}

```
