---
title: "Downloading and Visualizing Camels Data"
author: "Matthew Ross"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
<<<<<<< HEAD
library(broom)
=======
library(GGally)
>>>>>>> cc9960193fe9082e5d94cc2db4f3f6f2a143f480

knitr::opts_chunk$set(echo = TRUE)
```


# Data Acquisition

For this assignment we are going to be playing with annually
aggregated metrics of USGS flow data from the [CAMELS](https://ral.ucar.edu/solutions/products/camels) dataset. This dataset
has sparked a revolution in machine learning in hydrology. 


```{r}
if(!file.exists('data')){
  dir.create('data')
}

#climate means
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_clim.txt',
              'data/climate.txt')

#geologic controls
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_geol.txt',
              'data/geol.txt')

# soil data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_soil.txt',
              'data/soil.txt')

# topo data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_topo.txt',
              'data/topo.txt')


#vege 

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_vege.txt',
              'data/vege.txt')


#hydro (response variable)

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_hydro.txt',
              'data/hydro.txt')


# Variable definitions
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf',
              'data/meta.pdf')

```


## Data org


```{r}
dat_files <- list.files('data',
                        full.names = T)



climate <- read_delim(dat_files[1], delim = ';')

hydro <- read_delim('data/hydro.txt', delim = ';')
```

## Initial data viz


### Baseflow 


```{r}

<<<<<<< HEAD
all_vars <- inner_join(climate, hydro, by = 'gauge_id')

create_plot <- function(x, y){
  ggplot(all_vars, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point()+
    geom_smooth(method = "lm", se = F)
}

create_model <- function(x, y){
  model <- lm(all_vars[[y]] ~ all_vars[[x]], na.action = na.omit)
  
  
  output <- tidy(model)%>%
    filter(term != '(Intercept)')%>%
    mutate(x_term = x, 
           y_term = y, 
           r2 = summary(model)$r.squared, 
           sigma = summary(model)$sigma)%>%
    select(x_term, y_term, slope = estimate, p_value = p.value, r2, sigma)
  
  output

}
=======
ggplot(hydro, aes(x = baseflow_index,
                  y = q95)) + 
  geom_point()
>>>>>>> cc9960193fe9082e5d94cc2db4f3f6f2a143f480


```


Baseflow doesn't strongly control Q95 in a predictable way. 



### Climate controls


```{r}

cq <- inner_join(climate, hydro %>%
                   select(gauge_id, q95))



ggplot(cq, aes(x = p_mean, y = q95)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

p_mean_mod <- lm(q95 ~ p_mean, data = cq)

```


#### All at once

```{r}

png(filename = 'bigclimeplot.png', width = 10, height = 8, units = 'in', res = 300)
cq %>%
  select_if(is.numeric) %>%
  ggpairs()
dev.off()


long_cq <- cq %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = 'value',
               names_to = 'driver')

ggplot(long_cq, aes(value,
                    q95)) +
  geom_point() +
  facet_grid(~driver,
             scales = 'free')
```

The average precip (p_mean) controls 71% of the variation in 195, where every 1 mm/day increase in long-term average precip increases the q95 by 2.95 mm/day. 


# Assignment

## What are three controls on average runoff ratio? 

```{r}

#all column names other than runoff ratio
vars <- setdiff(names(all_vars%>% select_if(is.numeric)), c('gauge_id', 'runoff_ratio'))

plots <- map2(.x = "runoff_ratio", .y = vars , create_plot)
models <- map2(.x = "runoff_ratio", .y = vars , create_model)%>%
  bind_rows()%>%
  arrange(r2)

ggpubr::ggarrange(plotlist = plots)

ggplot(models, aes(x = y_term, y = r2))+
  geom_point()+
  #make x axis names vertical
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90))

# grab the models for q_mean, q95, aridity
models%>%
  filter(y_term %in% c('q95', 'low_prec_freq', 'q_mean'))
# only show the plots for these variables
vars <- c('q95', 'low_prec_freq', 'q_mean')
list_p <- map2(.x = "runoff_ratio", .y = vars , create_plot)


ggpubr::ggarrange(plotlist = list_p)
```

There are a few hydrological variables that are strongly correlated with runoff ratio. Mean Q is the most strongly correlated with runoff ratio, with a positive relationship (slope = 5.79) and predicts runoff ratio 76.7% of the time (pvalue = 3.07e-213). This makes sense since areas with higher flows are likely to have higher runoff created in their watersheds. 
A similar variable (q95) is positively correlated with runoff ratio and predicts runoff ratio  78.7% of the time. This shows that areas which experence higher peak flows, have higher runoff generation which is either indicative of the type of precip they recieve or the lands inability to take up that water and move it as baseflow. 

Low Precipitation frequency is negatively correlated with runoff ratio but is predicts runoff ratio less accurately(slope = -109, r2 = 0.527).Places where precipitation is very frequent will have more runoff generated as the soil is saturated and can't take up any more water, whereas places where rain is less frequent, the soil has time to absorb the water and less runoff is generated.


## What are three controls on baseflow_index? 

```{r}

#all column names other than runoff ratio
vars <- setdiff(names(all_vars%>% select_if(is.numeric)), c('gauge_id', 'baseflow_index'))

plots <- map2(.x = "baseflow_index", .y = vars , create_plot)
models <- map2(.x = "baseflow_index", .y = vars , create_model)%>%
  bind_rows()%>%
  arrange(r2)

ggpubr::ggarrange(plotlist = plots)

ggplot(models, aes(x = y_term, y = r2))+
  geom_point()+
  #make x axis names vertical
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90))



# grab the models for q_mean, q95, aridity
vars <- c('low_q_freq', 'high_q_freq', 'q5', 'aridity', 'low_q_dur', "stream_elas", 'low_prec_dur')

models%>%
  filter(y_term %in% vars)
# only show the plots for these variables

list_p <- map2(.x = "baseflow_index", .y = vars , create_plot)


ggpubr::ggarrange(plotlist = list_p)
```

Baseflow index seems to be primiarily controlled by low discharge frequency, s6



## What are three controls on mean flow? 




