---
title: "Downloading and Visualizing Camels Data"
author: "Matthew Ross"
date: "2024-04-22"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(GGally)

knitr::opts_chunk$set(echo = TRUE)
```

# Data Acquisition

For this assignment we are going to be playing with annually aggregated
metrics of USGS flow data from the
[CAMELS](https://ral.ucar.edu/solutions/products/camels) dataset. This
dataset has sparked a revolution in machine learning in hydrology.

```{r}
if(!file.exists('data')){
  dir.create('data')
}

#climate means
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_clim.txt',
              'data/climate.txt')

#geologic controls
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_geol.txt',
              'data/geol.txt')

# soil data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_soil.txt',
              'data/soil.txt')

# topo data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_topo.txt',
              'data/topo.txt')


#vege 

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_vege.txt',
              'data/vege.txt')


#hydro (response variable)

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_hydro.txt',
              'data/hydro.txt')


# Variable definitions
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf',
              'data/meta.pdf')

```

## Data org

```{r}
dat_files <- list.files('data',
                        full.names = T)



climate <- read_delim(dat_files[1], delim = ';')

hydro <- read_delim('data/hydro.txt', delim = ';')

geol <- read_delim('data/geol.txt', delim = ';')

soil <- read_delim('data/soil.txt', delim = ';')

topo <- read_delim('data/topo.txt', delim = ';')

vege <- read_delim('data/vege.txt', delim = ';')


```

## Initial data viz

<<<<<<< HEAD
### Baseflow
=======

### Baseflow 

>>>>>>> cc9960193fe9082e5d94cc2db4f3f6f2a143f480

```{r}
ggplot(hydro, aes(x = baseflow_index, y = q95))+
  geom_point()
```

### Climate Controls

```{r}
cq_1 <- inner_join(climate,hydro)

names(cq)

ggplot(cq, aes(x = p_mean, y = q95))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

p_mean_mod <- lm(q95 ~ p_mean, data = cq)

summary(p_mean_mod)

ggplot(hydro, aes(x = baseflow_index,
                  y = q95)) + 
  geom_point()


```

<<<<<<< HEAD
p_mean strongly controls q95

Average precip (p mean) controls 70% of the variation in q95, where
every 1 mm/day increase in average precip increased the q95 by 2.95 mm/d
=======

Baseflow doesn't strongly control Q95 in a predictable way. 



### Climate controls


```{r}

cq <- inner_join(climate, hydro %>%
                   select(gauge_id, q95))



ggplot(cq, aes(x = p_mean, y = q95)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F)

p_mean_mod <- lm(q95 ~ p_mean, data = cq)

```


#### All at once

```{r}

png(filename = 'bigclimeplot.png', width = 10, height = 8, units = 'in', res = 300)
cq %>%
  select_if(is.numeric) %>%
  ggpairs()
dev.off()


long_cq <- cq %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = 'value',
               names_to = 'driver')

ggplot(long_cq, aes(value,
                    q95)) +
  geom_point() +
  facet_grid(~driver,
             scales = 'free')
```

The average precip (p_mean) controls 71% of the variation in 195, where every 1 mm/day increase in long-term average precip increases the q95 by 2.95 mm/day. 

>>>>>>> cc9960193fe9082e5d94cc2db4f3f6f2a143f480

# Assignment

Baseflow doesnt strongly control q95 in a predicatble way

```{r}
long_cq <- cq %>% 
  select_if(is.numeric) %>% 
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = "value",
               names_to = "driver")
```

## What are three controls on average runoff ratio?

look at potential controls and only report the ones with high r-squared.
What we find here will be the basis of our variables when trying to
predict mean runoff

```{r}
cq_rr <- inner_join(climate,hydro %>% 
                   select(gauge_id, runoff_ratio))
```

```{r}
library(GGally)

png(filename = "bigrrplot.png", width = 12, height = 10, units = "in", res = 300)
cq_rr %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```

```{r}
ggplot(long_cq, aes(value, q95))+
  geom_point()+
  facet_grid(~driver, scales = "free")
```

```{r}

ggplot(cq_rr, aes(x = p_mean, y = runoff_ratio))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
pmean_runoffmod <- lm(runoff_ratio ~ p_mean, data = cq_rr)

summary(pmean_runoffmod)

```

With any one unit increase in p_mean, runoff_ratio can be expected to
increase by \~.117 A p-value of 2.2e-16 tells us this is a significant
relationship

###High Precipitation Frequency

```{r}

ggplot(cq_rr, aes(x = high_prec_freq, y = runoff_ratio))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
high_freq_runoffmod <- lm(runoff_ratio ~ high_prec_freq, data = cq_rr)

summary(high_freq_runoffmod)

```

With any one unit increase in high_prec_freq, runoff_ratio can be
expected to decrease by roughly -0.035 A p-value of 2.2e-16 tells us
this is a significant relationship

###Low Precipitation Frequency
```{r}

ggplot(cq_rr, aes(x = low_prec_freq, y = runoff_ratio))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
LowFreq_runoffmod <- lm(runoff_ratio ~ low_prec_freq, data = cq_rr)

summary(LowFreq_runoffmod)

```


With any one unit increase in low_prec_freq, runoff_ratio can be
expected to decrease by roughly -0.0048 A p-value of 2.2e-16 tells us 
this is a significant relationship


## What are three controls on baseflow_index?

frac_snow p_seasonality high_prec_freq

```{r}
cq_bf <- inner_join(climate,hydro %>% 
                   select(gauge_id, baseflow_index))

names(cq_bf)
```

```{r}
#library(GGally)

png(filename = "bigbaseflowplot.png", width = 10, height = 8, units = "in", res = 300)
cq_bf %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```

###Combining datasets to visualize more relationships
```{r}
gh <- inner_join(geol,hydro %>% 
                   select(gauge_id, baseflow_index))

names(gh)
```

```{r}
sh <- inner_join(soil,hydro %>% 
                   select(gauge_id, baseflow_index))

names(sh)
```

```{r}
th <- inner_join(topo,hydro %>% 
                   select(gauge_id, baseflow_index))

names(th)
```

```{r}
vh <- inner_join(vege,hydro %>% 
                   select(gauge_id, baseflow_index))

names(vh)
```

```{r}
#library(GGally)

png(filename = "bigbaseflowplotGH.png", width = 10, height = 8, units = "in", res = 300)
gh %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```

```{r}
#library(GGally)

png(filename = "bigbaseflowplotSH.png", width = 10, height = 8, units = "in", res = 300)
sh %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```

```{r}
#library(GGally)

png(filename = "bigbaseflowplotTH.png", width = 10, height = 8, units = "in", res = 300)
th %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```

```{r}
png(filename = "bigbaseflowplotVH.png", width = 10, height = 8, units = "in", res = 300)
vh %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```

###Slope_Mean
```{r}

ggplot(th, aes(x = slope_mean, y = baseflow_index))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
slope_baseflowmod <- lm(baseflow_index ~ slope_mean, data = th)

summary(slope_baseflowmod)

```

With any one unit increase in slope_mean, baseflow_index can be
expected to increase by ~0.0013 A p-value of < 2.2e-16 tells us 
this is a significant relationship


###Soil Conductivity
```{r}

ggplot(sh, aes(x = soil_conductivity, y = baseflow_index))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
soil_baseflowmod <- lm(baseflow_index ~ soil_conductivity, data = sh)

summary(soil_baseflowmod)

```


With any one unit increase in soil conductivity, baseflow_index can be
expected to increase by ~0.023 A p-value of 8.751e-9 tells us 
this is a significant relationship



### frac_snow
```{r}

ggplot(cq_bf, aes(x = frac_snow, y = baseflow_index))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
snow_baseflowmod <- lm(baseflow_index ~ frac_snow, data = cq_bf)

summary(snow_baseflowmod)

```


With any one unit increase in frac_snow, baseflow_index can be
expected to increase by ~0.312 A p-value of < 2.2e-16 tells us 
this is a significant relationship















## What are three controls on mean flow?

Combining datasets to visualize different factors

```{r}
gh_mf <- inner_join(geol,hydro %>% 
                   select(gauge_id, q_mean))

names(gh_mf)
```

```{r}
sh_mf <- inner_join(soil,hydro %>% 
                   select(gauge_id, q_mean))

names(sh_mf)
```

```{r}
th_mf <- inner_join(topo,hydro %>% 
                   select(gauge_id, q_mean))

names(th_mf)
```

```{r}
vh_mf <- inner_join(vege,hydro %>% 
                   select(gauge_id, q_mean))

names(vh_mf)
```

```{r}
cq_mf <- inner_join(climate,hydro %>% 
                   select(gauge_id, q_mean))
```

```{r}
#library(GGally)

png(filename = "bigmeanflowplotCH.png", width = 10, height = 8, units = "in", res = 300)
cq_mf %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```


```{r}
#library(GGally)

png(filename = "bigmeanflowplotTH.png", width = 10, height = 8, units = "in", res = 300)
th_mf %>% 
  select_if(is.numeric) %>% 
  ggpairs()
dev.off()
```

### p_mean
```{r}

ggplot(cq_mf, aes(x = p_mean, y = q_mean))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
pmean_meanflowmod <- lm(q_mean ~ p_mean, data = cq_mf)

summary(pmean_meanflowmod)

```


With any one unit increase in p_mean, q_mean can be
expected to increase by ~0.969 A p-value of < 2.2e-16 tells us 
this is a significant relationship


### low_prec_freq
```{r}

ggplot(cq_mf, aes(x = low_prec_freq, y = q_mean))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
lowprec_meanflowmod <- lm(q_mean ~ low_prec_freq, data = cq_mf)

summary(lowprec_meanflowmod)

```


With any one unit increase in low_prec_freq, q_mean can be
expected to decrease by ~0.031 A p-value of < 2.2e-16 tells us 
this is a significant relationship


### slope_mean
```{r}

ggplot(th_mf, aes(x = slope_mean, y = q_mean))+
  geom_point()+
  geom_smooth(method = "lm", se = F)

```

```{r}
slope_meanflowmod <- lm(q_mean ~ slope_mean, data = th_mf)

summary(slope_meanflowmod)

```


With any one unit increase in slope_mean, q_mean can be
expected to increase by ~0.017 A p-value of < 2.2e-16 tells us 
this is a significant relationship