---
title: "Downloading and Visualizing Camels Data"
author: "Matthew Ross"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)

install.packages("GGally")
library(GGally)
```

# Data Acquisition

For this assignment we are going to be playing with annually aggregated metrics of USGS flow data from the [CAMELS](https://ral.ucar.edu/solutions/products/camels) dataset. This dataset has sparked a revolution in machine learning in hydrology.

```{r}
if(!file.exists('data')){
  dir.create('data')
}

#climate means
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_clim.txt',
              'data/climate.txt')

#geologic controls
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_geol.txt',
              'data/geol.txt')

# soil data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_soil.txt',
              'data/soil.txt')

# topo data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_topo.txt',
              'data/topo.txt')


#vege 

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_vege.txt',
              'data/vege.txt')


#hydro (response variable)

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_hydro.txt',
              'data/hydro.txt')


# Variable definitions
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf',
              'data/meta.pdf')

```

## Data org

```{r}
dat_files <- list.files('data',
                        full.names = T)



climate <- read_delim(dat_files[1], delim = ';')

hydro <- read_delim('data/hydro.txt', delim = ';')

soil <- read_delim('data/soil.txt', delim = ';')

geol <- read_delim('data/geol.txt', delim = ';')
```

## Initial data viz

```{r}
#baseflow 
ggplot(hydro, aes(x=baseflow_index, y=q95))+
  geom_point()


#hydroclimate
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, baseflow_index))

ggplot(climate_q, aes(x=p_mean, y=q95))+
  geom_point()+
  geom_smooth(method = "lm", se=F)

p_mean_mod <- lm(q95~p_mean, data=climate_q)
summary(p_mean_mod)
```

the average precip controls 71% of the variation in 195, where every 1mm/day increase in long-term average precip increases the 9

\*r\^2 \> 0.3

correlation ^2^ = r\^2

# Assignment

## What are three controls on average runoff ratio?

```{r}
#climate runoff ratio join
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, runoff_ratio))

#climate x hydro runoff ratio correlatios plots
png(filename = "bigclimate.png", width = 10, height = 8, units = "in", res = 300)
climate_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#runoff ratio x aridity

ggplot(climate_q, aes(x=runoff_ratio, y=aridity))+
  geom_point()

#lm rr x aridity
climate_q_rr_mod <- lm(runoff_ratio~aridity, data=climate_q_rr) 
summary(climate_q_rr_mod) 

 
```

r\^2 = .327

```{r}
#runoff ratio x low_precip_freq
ggplot(climate_q, aes(x=runoff_ratio, y= low_prec_freq))+
  geom_point() 

#lm rr x low precip freq
climate_q_mod <- lm(runoff_ratio~low_prec_freq, data=climate_q) 
summary(climate_q_mod) 

```

r\^2 =.527

```{r}
#runoff ratio x high_precip_freq
ggplot(climate_q, aes(x=runoff_ratio, y= high_prec_freq))+
  geom_point() 

#lm rr x high precip freq
climate_q_mod <- lm(runoff_ratio~high_prec_freq, data=climate_q) 
summary(climate_q_mod) 
```

r\^2 = .462

## What are three controls on baseflow_index?

#frac_snow

#p_seasonality

#high_prec_freq

```{r}
#geol baseflow join
geol_q <- inner_join(geol, hydro %>%
                          select(gauge_id, baseflow_index))

#climate x hydro runoff ratio correlatios plots
png(filename = "biggeol.png", width = 10, height = 8, units = "in", res = 300)
geol_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#geol baseflow join
soil_q <- inner_join(soil, hydro %>%
                          select(gauge_id, baseflow_index))

#climate x hydro runoff ratio correlatios plots
png(filename = "bigsoil.png", width = 10, height = 8, units = "in", res = 300)
soil_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#climate runoff ratio join
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, baseflow_index))

#climate x hydro runoff ratio correlatios plots
png(filename = "bigclimate.png", width = 10, height = 8, units = "in", res = 300)
climate_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 

#baseflow x p_mean
ggplot(climate_q, aes(x=baseflow_index, y=high_prec_freq))+
  geom_point()

 climate_q_mod <- lm(baseflow_index~high_prec_freq, data=climate_q) 
summary(climate_q_mod) 
```

## What are three controls on mean flow?

#aridity

#pet_mean

#soil_conductivity

```{r}
climate_q <- inner_join(climate, hydro %>%
                          select(gauge_id, q_mean))

q_geol <- inner_join(geol, hydro %>%
                       select(gauge_id, q_mean))

soil_q <- inner_join(soil, hydro %>%
                       select(gauge_id, q_mean))

png(filename = "bigsoil.png", width = 10, height = 8, units = "in", res = 300)
soil_q %>%
  select_if(is.numeric)%>%
  ggpairs()
dev.off() 


```
