---
title: "Modelling Flow"
author: "Matthew Ross"
date: "2024-04-24"
output: html_document
---

```{r setup, include=FALSE}

install.packages('rpart.plot')
install.packages('yardstick')
library(tidyverse)
library(rpart)
library(rpart.plot)
library(yardstick)
library(randomForest)
knitr::opts_chunk$set(echo = TRUE)
```

# Modelling Flow

Now that we have explored individual correlations between long-term flow 
characteristics and potential drivers of those characteristics (climate,
soils, etc...), we can start to build out increasingly complex models
to predict flow characteristics. 



## Build a parsimonious linear model



```{r}

q_mean <- read_delim('data/hydro.txt', delim = ';') %>%
  select(gauge_id, q_mean) %>%
  inner_join(read_delim('data/climate.txt', delim = ';')) %>%
  inner_join(read_delim('data/soil.txt',delim = ';'))


q_mean_logs <- q_mean %>%
  mutate(p_log10 = log10(p_mean),
         aridlog10 = log10(aridity),
         q_mean10 = log10(q_mean))

naive_mod <- lm(q_mean10 ~ aridlog10 * p_log10, data = q_mean_logs)

summary(naive_mod)


```


## Build a CART model to predict flow. 
 

```{r}
set.seed(22157)
q_4 <- q_mean_logs %>%
  select(q_mean10, p_log10, aridlog10, soil_depth_pelletier, max_water_content, organic_frac, frac_snow,pet_mean) %>%
  mutate(q_class = cut_number(q_mean10,n = 4)) %>%
  select(-q_mean10) %>%
  na.omit()

train <- q_4 %>%
  sample_frac(.7)

test <- q_4 %>%
  anti_join(train)

cart_simple <- rpart(q_class ~., data = train, cp = 0.01,
                     method = 'class')


plot(cart_simple)
text(cart_simple, cex = 0.8, use.n = TRUE, xpd = TRUE)


test$pred <- predict(cart_simple, test, 'class')
cm <- conf_mat(test, q_class,pred)

autoplot(cm, type = "heatmap") +
  scale_fill_gradient(low="#D6EAF8",high = "#2E86C1")

accuracy(test,q_class,pred)
```


## Build a RandomForest



### Classification Forest

```{r}

summary(train)
rf_class <- randomForest(q_class ~ ., data = train, 
                         maxnodes = 9, 
                         nPerm = 2,
                         mtry = 5,
                         importance = T)

test$rf_pred <- predict(rf_class, test)

cm_rf <- conf_mat(test, q_class,rf_pred)

autoplot(cm_rf, type = "heatmap") +
  scale_fill_gradient(low="#D6EAF8",high = "#2E86C1")

accuracy(test,q_class,rf_pred)

```

### Continuous pred

```{r}
q_rf <- q_mean_logs %>%
  select(q_mean10, p_log10, aridlog10, soil_depth_pelletier,  frac_snow,pet_mean) %>%
  na.omit()

train_cont <- q_rf %>%
  sample_frac(.7)

test_cont <- q_rf %>%
  anti_join(train)


rf_numer <- randomForest(q_mean10 ~ ., data = train_cont, 
                         maxnodes = 50, 
                         nPerm = 5,
                         mtry = 8,
                         importance = T)

rf_numer$importance

test_cont$rf_pred <- predict(rf_numer, test_cont)
train_cont$rf_pred <- predict(rf_numer, train_cont)


ggplot(test_cont, aes(x = q_mean10,
                      y = rf_pred)) + 
  geom_point() + 
  geom_abline(slope = 1)


cor(test_cont$rf_pred,test_cont$q_mean10)^2
cor(train_cont$rf_pred,train_cont$q_mean10)^2

```


##Assignment


```{r}

runoff_ratio <- read_delim('data/hydro.txt', delim = ';') %>%
  select(gauge_id, runoff_ratio) %>%
  inner_join(read_delim('data/climate.txt', delim = ';')) %>%
  inner_join(read_delim('data/soil.txt',delim = ';'))


runoff_ratio_logs <- runoff_ratio %>%
  mutate(p_log10 = log10(p_mean),
         aridlog10 = log10(aridity),
         runoff_ratio_10 = log10(runoff_ratio))

runoff_ratio_mod <- lm(runoff_ratio_10 ~ aridlog10 * p_log10, data = runoff_ratio_logs)


summary(runoff_ratio_mod)

```


## Build a CART model to predict flow. 



```{r}
set.seed(22157)
runoff_ratio_4 <- runoff_ratio_logs %>%
  select(runoff_ratio_10, p_log10, aridlog10, soil_depth_pelletier, max_water_content, organic_frac, frac_snow,pet_mean) %>%
  mutate(runoff_ratio_class = cut_number(runoff_ratio_10,n = 4)) %>%
  select(-runoff_ratio_10) %>%
  na.omit()

train <- runoff_ratio_4 %>%
  sample_frac(.7)

test <- runoff_ratio_4 %>%
  anti_join(train)

cart_simple <- rpart(runoff_ratio_class ~., data = train, cp = 0.01,
                     method = 'class')


plot(cart_simple)
text(cart_simple, cex = 0.8, use.n = TRUE, xpd = TRUE)


test$pred <- predict(cart_simple, test, 'class')
cm <- conf_mat(test, runoff_ratio_class,pred)

autoplot(cm, type = "heatmap") +
  scale_fill_gradient(low="#D6EAF8",high = "#2E86C1")

accuracy(test,runoff_ratio_class,pred)
```




### Classification Forest

```{r}

summary(train)
rf_class <- randomForest(runoff_ratio_class ~ ., data = train, 
                         maxnodes = 9, 
                         nPerm = 2,
                         mtry = 5,
                         importance = T)

test$rf_pred <- predict(rf_class, test)

cm_rf <- conf_mat(test, runoff_ratio_class,rf_pred)

autoplot(cm_rf, type = "heatmap") +
  scale_fill_gradient(low="#D6EAF8",high = "#2E86C1")

accuracy(test,runoff_ratio_class,rf_pred)

```

### Continuous pred

```{r}
runoff_ratio_rf <- runoff_ratio_logs %>%
  select(runoff_ratio_10, p_log10, aridlog10, soil_depth_pelletier,  frac_snow,pet_mean) %>%
  na.omit()

train_cont <- runoff_ratio_rf %>%
  sample_frac(.7)

test_cont <- runoff_ratio_rf %>%
  anti_join(train)


rf_numer <- randomForest(runoff_ratio_10 ~ ., data = train_cont, 
                         maxnodes = 50, 
                         nPerm = 5,
                         mtry = 8,
                         importance = T)

rf_numer$importance

test_cont$rf_pred <- predict(rf_numer, test_cont)
train_cont$rf_pred <- predict(rf_numer, train_cont)


ggplot(test_cont, aes(x = runoff_ratio_10,
                      y = rf_pred)) + 
  geom_point() + 
  geom_abline(slope = 1)


cor(test_cont$rf_pred,test_cont$runoff_ratio_10)^2
cor(train_cont$rf_pred,train_cont$runoff_ratio_10)^2

```