---
title: "Modelling Flow"
author: "Matthew Ross"
date: "2024-04-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Modelling Flow

Now that we have explored individual correlations between long-term flow
characteristics and potential drivers of those characteristics (climate,
soils, etc...), we can start to build out increasingly complex models to
predict flow characteristics.

# Assignment

## Build a parsimonious linear model

Pick one of the flow characteristics that mosts interests you and use
the `lm` function to build a parsimonious model that predicts your
favorite characteristic. What is parsimony? It's a model that is complex
enough to capture the major controls, but simple enough to be
understandable. You will need to explore, read, understand the
differences between a + sign interaction, a ":" interaction and a \*
interaction in `lm` terminology.

Please report your R2, slope estimates, and p-value of your model and
write out your interpretation of these numbers.

```{r}
#plot simple relationship
#create parsimonious model
slope_mod <- lm(q_mean ~ slope_fdc * runoff_ratio * baseflow_index, data = hydro)

# Get the summary of the model
summary(slope_mod)
```

**Answer:**

Slope Estimate:

-   A one-unit increase in **`slope_fdc`** is associated with a decrease
    of \~0.5330 units in **`q_mean`**

-   A one-unit increase in **`runoff_ratio`** is associated with an
    increase of \~0.9667 units in **`q_mean`**.

-   A one-unit increase in **`baseflow_index`** is associated with a
    decrease of \~0.3779 units in **`q_mean`**

r-squared = 0.8056

p-value = \< 2e-16

## Build a CART model to predict flow.

Linear models help us both predict and understand drivers of change,
machine learning can help us understand drivers of change, but as a
technique it is more suited to accurate predictions. CART or
Classification and Regression Trees are a nice intermediate between lms
and ml. Tons of resources for this but [CART
Logic](https://koalaverse.github.io/machine-learning-in-R/decision-trees.html#cart-software-in-r),
provides a good conceptual overview, and [CART
demo](https://www.statmethods.net/advstats/cart.html) provides a good
enough code demo.

Read the logic intro above, and the code demo as well, to build a CART
model version of your lm. Use the code to visualize your CART output.

```{r}
library(rpart)

cart_mod <- rpart(q_mean ~ slope_fdc+runoff_ratio+baseflow_index, data=hydro, method = "anova")

# visualize cross-validation results
plotcp(cart_mod)

#summarize data
summary(cart_mod)

# create additional plots
rsq.rpart(cart_mod) # visualize cross-validation results

# plot tree
plot(cart_mod, uniform=TRUE,
   main="Regression Tree for Mean Flow ")
text(cart_mod, use.n=TRUE, all=TRUE, cex=.8)
##not sure that this shows much information... did I do it wrong?
```

## Build a RandomForest

CARTs are a single tree, what if we had thousands? Would we get better
performance (yes!)

The same CART logic site above introduces random forests as well. Please
read this part of the site and use the code demo to build your own
RandomForest. Remember, for a RandomForest type model we want to make
sure we split our data at least into train and test datasets and ideally
into train-test-val.

```{r}

```
