---
title: "Modelling Flow"
author: "Matthew Ross"
date: "2024-04-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

library(tidyverse)
library(rpart)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE)
```

# Modelling Flow

Now that we have explored individual correlations between long-term flow
characteristics and potential drivers of those characteristics (climate,
soils, etc...), we can start to build out increasingly complex models to
predict flow characteristics.

# Assignment

## Build a parsimonious linear model

Pick one of the flow characteristics that mosts interests you and use
the `lm` function to build a parsimonious model that predicts your
favorite characteristic. What is parsimony? It's a model that is complex
enough to capture the major controls, but simple enough to be
understandable. You will need to explore, read, understand the
differences between a + sign interaction, a ":" interaction and a \*
interaction in `lm` terminology.

Please report your R2, slope estimates, and p-value of your model and
write out your interpretation of these numbers.

```{r}
#create a dataframe with combining hydro and climate
chydro <- inner_join(hydro,climate)

#plot simple relationship
#create parsimonious model
q_mod <- lm(q_mean ~ runoff_ratio * p_mean, data = chydro)

#Get the summary of the model
summary(q_mod)

q_mean_logs <- chydro %>%
  mutate(q_log10=log10(q_mean), runofflog10 = log10(runoff_ratio))

q_mod_log <- lm(q_log10 ~ runofflog10 * p_mean, data = chydro)

ggplot(chydro, aes(x = q_mean, y = runoff_ratio, color = p_mean)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = 'lm')
```

```{r}

#How Matt did it in class

q_mean <- read_delim('data/hydro.txt', delim = ';') %>%
  select(gauge_id, q_mean) %>%
  inner_join(read_delim('data/climate.txt', delim = ';')) %>%
  inner_join(read_delim('data/soil.txt', delim = ';'))

#building linear model, predicting q_mean as a function of p_mean

names(q_mean)

##this is testing to see if low_prec_timing influences the q_mean and p_mean... 
naive_mod <- lm(q_mean ~ p_mean * low_prec_timing, data = q_mean)

summary(naive_mod)

##the above model shows us a lot of intereactions that are NOT significant, so you would not want to use this

##if you change the * to + -> you get a change in intercept but not a change in slope

##intercept means the average q_mean without any precip, water_frac or aridity -> interpretation get's increasingly ghard when you add an interpretation variable (* in lm formula). 

##this complicated interaction is why we may want to switch to machine learning... 

#logged data
q_mean_logs <- q_mean %>%
  mutate(p_log10=log10(p_mean), aridlog10 = log10(aridity),
         q_mean10 = log10(q_mean))

naive_mod <- lm(q_mean10 ~ aridlog10 * p_log10, data = q_mean_logs)

ggplot(q_mean, aes(x = p_mean, y = q_mean, color = aridity)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = 'lm')

##this model is working better, but there is still some variation that we are not explaining... 

#to improve the model, check the indicators ands ee what relationship exists
```

**Answer:**

-   Slope Estimates:

    -   Intercept: 2.873e-04

    -   runoff_ratio: -1.530e-03

    -   p_mean: -9.969e-05

    -   runoff_ratio:p_mean: 1.000e+00

-   R-squared value:

    -   Multiple R-squared: 1

    -   Adjusted R-squared: 1

-   P-values:

    -   Intercept: 0.4034

    -   runoff_ratio: 0.0358

    -   p_mean: 0.4274

    -   runoff_ratio:p_mean: \< 2.2e-16

-   Interpretation:

## Build a CART model to predict flow.

Linear models help us both predict and understand drivers of change,
machine learning can help us understand drivers of change, but as a
technique it is more suited to accurate predictions. CART or
Classification and Regression Trees are a nice intermediate between lms
and ml. Tons of resources for this but [CART
Logic](https://koalaverse.github.io/machine-learning-in-R/decision-trees.html#cart-software-in-r),
provides a good conceptual overview, and [CART
demo](https://www.statmethods.net/advstats/cart.html) provides a good
enough code demo.

Read the logic intro above, and the code demo as well, to build a CART
model version of your lm. Use the code to visualize your CART output.

```{r}
#install.packages('yardstick')
#install.packages('rpart.plot')
library(yardstick)
library(rpart.plot)

#matt's example
q_4 <- q_mean_logs %>%
  select(q_mean10, p_log10, aridlog10, sand_frac, clay_frac, soil_depth_pelletier) %>%
  #to make it a classification tree - converting data from continuous to a class (binning it) -> gives us equal bins in each classification, which is necessary for machine learning
  mutate(q_class = cut_number(q_mean10, n=4)) %>%
  select(-q_mean10)

train <- q_4 %>%
  sample_frac(.7)

test <- q_4 %>%
  anti_join(train)

cart_simple <- rpart(q_class ~ ., data = train, cp = 0.001, method = 'class')

#visualize table/classes  
table(q_4$q_class)

#cart_simple <- rpart(q_class ~., data = q_4, cp = 0.01, method = 'anova')

plot(cart_simple)
text(cart_simple, cex = 0.8, use.n = TRUE, xpd = TRUE)

#how to understand model's performance - comparing predictions with actual data
q_4$pred <- predict(cart_simple, q_4, 'class')
cm <- conf_mat(q_4, q_class, pred)

autoplot(cm, type = 'heatmap') +
  scale_fill_gradient(low = 'grey', high = 'blue')

accuracy(test_d, group, guess)

test$pred <- predict(cart_simple, test, 'class')
cm <- conf_mat(q_4, q_class, pred)
```

```{r}
##need to bin my data using cart_numbers step into 4ish? ... change it to using the chydro dataset, and variables from only the climate, with q_mean from hydro!
#my cart model
cart_mod <- rpart(q_mean ~ slope_fdc+runoff_ratio+baseflow_index, data=hydro, method = "anova")

# visualize cross-validation results
plotcp(cart_mod)

#summarize data
summary(cart_mod)

# create additional plots
rsq.rpart(cart_mod) # visualize cross-validation results

# plot tree
plot(cart_mod, uniform=TRUE,
   main="Regression Tree for Mean Flow ")

text(cart_mod, cex = 0.8, use.n = TRUE, xpd = TRUE)
```

## Build a RandomForest

CARTs are a single tree, what if we had thousands? Would we get better
performance (yes!)

The same CART logic site above introduces random forests as well. Please
read this part of the site and use the code demo to build your own
RandomForest. Remember, for a RandomForest type model we want to make
sure we split our data at least into train and test datasets and ideally
into train-test-val.

```{r}

```
