---
title: "Downloading and Visualizing Camels Data"
author: "Matthew Ross"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(echo = TRUE)
```

# Data Acquisition

For this assignment we are going to be playing with annually aggregated metrics of USGS flow data from the [CAMELS](https://ral.ucar.edu/solutions/products/camels) dataset. This dataset has sparked a revolution in machine learning in hydrology.

```{r}
if(!file.exists('data')){
  dir.create('data')
}

#climate means
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_clim.txt',
              'data/climate.txt')

#geologic controls
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_geol.txt',
              'data/geol.txt')

# soil data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_soil.txt',
              'data/soil.txt')

# topo data
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_topo.txt',
              'data/topo.txt')


#vege 

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_vege.txt',
              'data/vege.txt')


#hydro (response variable)

download.file('https://gdex.ucar.edu/dataset/camels/file/camels_hydro.txt',
              'data/hydro.txt')


# Variable definitions
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf',
              'data/meta.pdf')

```

## Data org

```{r}
dat_files <- list.files('data',
                        full.names = T)



climate <- read_delim(dat_files[1], delim = ';')

hydro <- read_delim('data/hydro.txt', delim = ';')
```

## Initial data viz

Baseflow

```{r}
# this is what Matt would do to start w/ climate data

# predicting high flow but not max flow

ggplot(hydro, aes(x=baseflow_index, y = q95)) +
  geom_point()
# base flow may predict 95th quantiles?
# in this dataset,low baseflow = low long term runoff
# this variable doesn't look like it's highly predictive...

```

Hydroclimate

```{r}

cq <- inner_join(climate, hydro %>%
  select(gauge_id, q95)) # joining by gauge ID

names(cq)

ggplot(cq, aes(x = p_mean, y = q95))+
  geom_point() +
geom_smooth(method = "lm", se = F)
# there is agreement here!
# P_means strongly controls q95

# we want to get a number to describe this relationship
p_mean_mod <- lm(q95 ~ p_mean, data = cq)
summary(p_mean_mod)

# the average precip (p_mean) controls 71% of the variation in q95, where every 1 mm/day increase in longterm average precip increases the q95 by 2.95 mm/day. This is likely bc more water means this landscape gets wetter etc...
```

# Assignment

look at potential controls and only report those with a relatively high R squared - like higher than 0.3. what we find here will help us with our predictions later.

```{r}
# reading in rest of the data that I anticipate using:

geol <- read_delim('data/geol.txt', delim = ';')
soils <- read_delim('data/soil.txt', delim = ';')
topo <- read_delim('data/topo.txt', delim = ';')
vege <- read_delim('data/vege.txt', delim = ';')
```

## What are three controls on average runoff ratio?

```{r}
# prepping datasets to work with

cq_runoff <- inner_join(climate, hydro %>%
  select(gauge_id, runoff_ratio))

names(cq_runoff)

topo_runoff <- inner_join(topo, hydro %>%
  select(gauge_id, runoff_ratio))

names(topo_runoff)


clim_runoff <- inner_join(climate, hydro %>%
  select(gauge_id, runoff_ratio))

names(clim_runoff)
```

aridity

```{r}
ggplot(cq_runoff, aes(x = aridity, y = runoff_ratio))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
aridity_mod <- lm(runoff_ratio ~ aridity, data = cq_runoff)
summary(aridity_mod)
```

Aridity controls 32% of the variation in runoff_ratio, where every increase of 1 in aridity decreases runoff_ratio by -0.217.

slope_mean

```{r}
ggplot(topo_runoff, aes(x = slope_mean, y = runoff_ratio))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
aridity_mod <- lm(runoff_ratio ~ slope_mean, data = topo_runoff)
summary(aridity_mod)
```

Slope controls 38% of the variation in runoff_ratio, where every increase of 1 m/km of slope over the catchment area increases runoff_ratio by 0.00305.

p_mean

```{r}
ggplot(clim_runoff, aes(x = p_mean, y = runoff_ratio))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
pmean_mod <- lm(runoff_ratio ~ p_mean, data = clim_runoff)
summary(pmean_mod)
```

P_mean controls 50% of the variation in runoff_ratio, where every increase of 1 mm/day increases runoff_ratio by 0.117.

## What are three controls on baseflow_index?

```{r}

# getting datasets prepared to work with:

soilq <- inner_join(soils, hydro %>%
  select(gauge_id, baseflow_index))

names(soilq)

geolq <- inner_join(geol, hydro %>%
  select(gauge_id, baseflow_index)) 

names(geolq)

cq2 <- inner_join(climate, hydro %>%
  select(gauge_id, baseflow_index))

names(cq2)

topoq <- inner_join(topo, hydro %>%
  select(gauge_id, baseflow_index))

names(topoq)

vegeq <- inner_join(vege, hydro %>%
  select(gauge_id, baseflow_index))

names(vegeq)

```

high_q_freq

```{r}
ggplot(hydro, aes(x = high_q_freq, y = baseflow_index))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
highq_mod <- lm(baseflow_index ~ high_q_freq, data = hydro)
summary(highq_mod)
```

The frequency of high flow days controls 39% of the variation in baseflow_index, where every 1 day/year increase in the frequency of high flow days decreases baseflow_index by -0.0035.

low_q_freq

```{r}
ggplot(hydro, aes(x = low_q_freq, y = baseflow_index))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
lowq_mod <- lm(baseflow_index ~ low_q_freq, data = hydro)
summary(lowq_mod)
```

The frequency of low flow days controls 73% of the variation in baseflow_index, where every 1 day/year increase in the frequency of high flow days decreases baseflow_index by -1.697e-03.

q5

```{r}
ggplot(hydro, aes(x = q5, y = baseflow_index))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
q5_mod <- lm(baseflow_index ~ q5, data = hydro)
summary(q5_mod)
```

Q5 controls 24% of the variation in baseflow_index, where every 1 mm/day increase in the frequency of high flow days increases baseflow_index by 0.295.

## What are three controls on mean flow?

```{r}
# prepping datasets to work with:

clim_flow <- inner_join(climate, hydro %>%
  select(gauge_id, q_mean))
```

p_mean

```{r}
ggplot(clim_flow, aes(x = p_mean, y = q_mean))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
pmean_flow_mod <- lm(q_mean ~ p_mean, data = clim_flow)
summary(pmean_flow_mod)
```

P_mean controls 78% of the variation in mean flow, where every 1 mm/day increase in p_mean increases mean flow by 0.969.

runoff_ratio

```{r}
ggplot(hydro, aes(x = runoff_ratio, y = q_mean))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
runoff_flow_mod <- lm(q_mean ~ runoff_ratio, data = hydro)
summary(runoff_flow_mod)
```

Runoff_ratio controls 76% of the variation in mean flow, where every increase of 1 in runoff_ratio increases mean flow by 5.792.

q95

```{r}
ggplot(hydro, aes(x = q95, y = q_mean))+
  geom_point() +
geom_smooth(method = "lm", se = F)

# we want to get a number to describe this relationship
q95_flow_mod <- lm(q_mean ~ q95, data = hydro)
summary(q95_flow_mod)
```

Q95 controls 92% of the variation in mean flow, where every increase of 1 in q95 increases mean flow by 0.301.

===

Following doesn't pertain to the assignment - it's just notes from class.

Notes

-   springs don't care how much rain there was, their discharge is usually always the same

-   sand/silt/clay, hydrographs/baseflow,

==

```{r}
library(GGally)
# using easier way to look at correlations between everything:
png(filename = "bigclimateplot.png", width = 10, height = 8, units = "in", res = 300)
cq %>%
  select_if(is.numeric) %>%
  ggpairs() %>%
  dev.off()

long_cq <- cq %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = "value",
               names_to = "driver")

ggplot(long_cq, aes(value, q95)) +
  geom_point() +
  facet_grid(~driver, scales = "free")

###########

long_climrun <- clim_runoff %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = "value",
               names_to = "driver")

ggplot(long_climrun, aes(value, runoff_ratio)) +
  geom_point() +
  facet_grid(~driver, scales = "free")

##############

long_clim_flow <- clim_flow %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = p_mean:low_prec_dur,
               values_to = "value",
               names_to = "driver")

ggplot(long_clim_flow, aes(value, q_mean)) +
  geom_point() +
  facet_grid(~driver, scales = "free")

##########3


long_hydro <- hydro %>%
  select_if(is.numeric) %>%
  pivot_longer(cols = runoff_ratio:hfd_mean,
               values_to = "value",
               names_to = "driver")

ggplot(long_hydro, aes(value, q_mean)) +
  geom_point() +
  facet_grid(~driver, scales = "free")
```
