---
title: "Modelling Flow"
author: "Matthew Ross"
date: "2024-04-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Modelling Flow

Now that we have explored individual correlations between long-term flow
characteristics and potential drivers of those characteristics (climate,
soils, etc...), we can start to build out increasingly complex models to
predict flow characteristics.

# Assignment

## Build a parsimonious linear model

Pick one of the flow characteristics that mosts interests you and use
the `lm` function to build a parsimonious model that predicts your
favorite characteristic. What is parsimony? It's a model that is complex
enough to capture the major controls, but simple enough to be
understandable. You will need to explore, read, understand the
differences between a + sign interaction, a ":" interaction and a \*
interaction in `lm` terminology.

Please report your R2, slope estimates, and p-value of your model and
write out your interpretation of these numbers.

\*\*I don't have a ton of experience in MLR analyses and I would love to
learn more with examples in class on Monday!

**Looking at mean flow:**

```{r}
# putting together dataframe:
mlr_climq <- inner_join(hydro, climate %>%
  select(gauge_id, p_mean, ))

# runoff ratio, q95, and p_mean as controls on mean flow:
mlr_model <- lm(formula = q_mean ~ runoff_ratio + q95 + p_mean, data = mlr_climq)

summary(mlr_model)
```

**Interpretation of model results:**

-   R\^2 value = 0.9515

-   p value:

    -   runoff_ratio = 6.07e-15

    -   q95 = \<2e-16

    -   p_mean = \<2e-16

-   slopes:

    -   runoff_ratio = 0.988873

    -   q95 = 0.186571

    -   p_mean = 0.302800

**Interpretation:**

The R\^2 value for this model is high at 0.9515, which means that
approximately 95.15% of the variance in the dependend variable is
explained by the independent variables. All 3 predictor variables have
very small p-values, suggesting a highly significant relationship
between all of the independent/predictor variables (runoff ratio, high
flow, and mean daily precipitation) and the dependent variable of mean
flow.

## Build a CART model to predict flow.

Linear models help us both predict and understand drivers of change,
machine learning can help us understand drivers of change, but as a
technique it is more suited to accurate predictions. CART or
Classification and Regression Trees are a nice intermediate between lms
and ml. Tons of resources for this but [CART
Logic](https://koalaverse.github.io/machine-learning-in-R/decision-trees.html#cart-software-in-r),
provides a good conceptual overview, and [CART
demo](https://www.statmethods.net/advstats/cart.html) provides a good
enough code demo.

Read the logic intro above, and the code demo as well, to build a CART
model version of your lm. Use the code to visualize your CART output.

```{r}
library(rpart)

cart_model <- rpart(formula = q_mean ~ runoff_ratio + q95 + p_mean, method = "anova", data = mlr_climq)

summary(cart_model) # I do not understand the output of this model...

# I got the following code for visualization from the CART Demo link:

# visualizing cart model:
plotcp(cart_model) # I also do not understand how to interpret this output
# but it seems like the "Inf" value in the first tree isn't great?

# creating additional plots
par(mfrow=c(1,2)) # two plots on one page
rsq.rpart(cart_model) # visualize cross-validation results


# plotting the tree
plot(cart_model, uniform=TRUE,
   main="Regression Tree for Mean Flow")
text(cart_model, use.n=TRUE, all=TRUE, cex=.8) # for some reason the top text/part of the tree gets cut off, not sure how to fix this with the base level plot/text functions

# also don't really know how to interpret this - I don't have a ton of experience in stats, regrettably


```

\

## Build a RandomForest

CARTs are a single tree, what if we had thousands? Would we get better
performance (yes!)

The same CART logic site above introduces random forests as well. Please
read this part of the site and use the code demo to build your own
RandomForest. Remember, for a RandomForest type model we want to make
sure we split our data at least into train and test datasets and ideally
into train-test-val.

```{r}
library(randomForest)

# need to split the mlr_clim dataset into train and test, but I'm not exactly sure how to do that in an effective way

# What is train-test-val?

# I was thinking that I would use the randomForest package for my analysis, as that seemed to be the most commonly used package for building a randomforest model. I looked at the example code and I wasn't quite able to figure out how to put it into practice with my own code/MLR example. I think that seeing a live example/explanation of the code would be really helpful for me, that's the way that I've found I'm able to grasp concepts the best.

```
